---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MethylSurroGetR

<!-- badges: start -->
[![R-CMD-check](https://github.com/jagoode27/MethylSurroGetR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jagoode27/MethylSurroGetR/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: GPL (>= 3)](https://img.shields.io/badge/License-GPL%20%28%3E%3D%203%29-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![Codecov test coverage](https://codecov.io/gh/jagoode27/MethylSurroGetR/graph/badge.svg)](https://app.codecov.io/gh/jagoode27/MethylSurroGetR)
<!-- badges: end -->

### 🚧 Under Development 🚧
**This package is still in development and not yet ready for general use. Proceed with caution!**

The goal of MethylSurroGetR is to offer user-friendly tools and functions for generating predicted values from existing DNA methylation surrogates. It aims to enhance the accessibility and reproducibility of data analysis with methylation surrogates.

## Installation

You can install the development version of MethylSurroGetR from GitHub with:

``` r
# install.packages("remotes")
remotes::install_github("jagoode27/MethylSurroGetR")
```

## Example

This is a basic example workflow:

```{r example, eval=FALSE}
library(MethylSurroGetR)

# Load example data
data(beta_matrix_comp)
data(wts_df)
data(ref_df)

# Create weights vector
wts_vec <- setNames(wts_df$wt_lin, rownames(wts_df))

# Create surrogate object
my_surro <- surro_set(
  methyl = beta_matrix_comp,
  weights = wts_vec,
  intercept = "Intercept"
)

# Fill missing probes with reference values
ref_vec <- setNames(ref_df$mean, rownames(ref_df))
my_surro <- reference_fill(my_surro, reference = ref_vec)

# Calculate predictions
predictions <- surro_calc(my_surro, transform = "linear")
```

## Getting Help

See the [Getting Started vignette](https://jagoode27.github.io/MethylSurroGetR/articles/MethylSurroGetR.html) for detailed examples and workflows.

For bug reports and feature requests, please [open an issue](https://github.com/jagoode27/MethylSurroGetR/issues).
