---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MethylSurroGetR

<!-- badges: start -->
[![R-CMD-check](https://github.com/jagoode27/MethylSurroGetR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jagoode27/MethylSurroGetR/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: GPL (>= 3)](https://img.shields.io/badge/License-GPL%20%28%3E%3D%203%29-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![Codecov test coverage](https://codecov.io/gh/jagoode27/MethylSurroGetR/graph/badge.svg)](https://app.codecov.io/gh/jagoode27/MethylSurroGetR)
<!-- badges: end -->

### 🚧 Under Development 🚧
**This package is still in development and not yet ready for general use. Proceed with caution!**

MethylSurroGetR is an R package designed to facilitate the construction of DNA methylation surrogate biomarker values from existing studies. It provides comprehensive tools for handling missing methylation data through multiple imputation strategies, including mean/median imputation for missing observations and reference-based filling for missing probes. The package supports flexible surrogate calculation with linear, logistic, and Poisson transformations, comprehensive input validation, and detailed diagnostic reporting. Additional utilities include conversion between beta values and M-values, as well as missing data analysis.

## Installation

You can install the development version of MethylSurroGetR from GitHub with:

``` r
# install.packages("remotes")
remotes::install_github("jagoode27/MethylSurroGetR")
```

## Example

This is a basic example workflow:

```{r example, eval=FALSE}
library(MethylSurroGetR)

# Load example data
data(beta_matrix_comp)
data(wts_df)
data(ref_df)

# Create weights vector
wts_vec <- setNames(wts_df$wt_lin, rownames(wts_df))

# Create surrogate object
my_surro <- surro_set(
  methyl = beta_matrix_comp,
  weights = wts_vec,
  intercept = "Intercept"
)

# Fill missing probes with reference values
ref_vec <- setNames(ref_df$mean, rownames(ref_df))
my_surro <- reference_fill(my_surro, reference = ref_vec)

# Calculate predictions
predictions <- surro_calc(my_surro, transform = "linear")
```

## Getting Help

See the [Getting Started vignette](https://jagoode27.github.io/MethylSurroGetR/articles/MethylSurroGetR.html) for detailed examples and workflows.

For bug reports and feature requests, please [open an issue](https://github.com/jagoode27/MethylSurroGetR/issues).
