---
title: "MethylSurroGetR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MethylSurroGetR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(tidyverse)
```

# Installing & Loading Package

We can use the `devtools` package to install the development version of `MethylSurroGetR` from GitHub.

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("jagoode27/MethylSurroGetR")
```

Once the package is installed, we can load the package.
```{r}
library(MethylSurroGetR)
```

# Loading Sample Data

At a minimum, `MethylSurroGetR` requires two input objects: a methylation matrix and a named vector, matrix, or data frame with surrogate weights.

## Methylation Matrix

We'll start by loading our matrix of beta values.

```{r}
data("beta_matrix_miss")
print(beta_matrix_miss)
```

Notice that probe IDs are in row names and sample IDs are in column names. This is the required orientation in `MethylSurroGetR`.

## Surrogate Weights

Next, we'll load a named vector of weights to construct the surrogate.

```{r}
data("wts_vec_lin")
print(wts_vec_lin)
```

# Building `mehtyl_surro` Data Object

In order to generate predicted values from surrogate weights, we need to construct a `mehtyl_surro` object. To do this, we will use the `surro_set()` function and specify the methylation matrix with the `methyl` and the weights object with the `weights` option. In addition, we need to specify the character string for the name of the intercept in the weights object (if applicable) with the `intercept` option.

```{r}
my_surrogate <- surro_set(methyl = beta_matrix_miss,
                          weights = wts_vec_lin,
                          intercept = "Intercept")
print(my_surrogate)
```

We can see that the resulting `methyl_surro` object has three elements that correspond to the inputs: `methyl`, `weights`, and `intercept`. In addition to placing each of the inputs into their own object element, the function has also subset the methylation matrix to include all of the probes included in the weights (and nothing more). Although it makes little difference with this toy example, methylation matrices in the real world contain hundeds of thousands of probes, while many surrogates are constructed using tens of thousands of probes.

# Handling Missing Values

Notice that there are a number of missing values in the `methyl` element of `my_surrogate`. Our next step is to address these.

## Two Types of Missing Values

`MethylSurroGetR` distinguishes between two types of missing values:

  - <font color='blue'><b>Missing Observatons:</b></font> These are probes that were measured in the target data, with missing values for some samples.
    - Notice that `cg01` was missing in our methylation matrix for `samp2`, `samp3`, and `samp5`. Similarly, `cg07` was missing for `samp5`.
  - <font color='red'><b>Missing Probes:</b></font> These are probes that are not measured in the target data. Therefore, they are missing for all samples in the methylation matrix. This might have occurred because those probes was removed during QC or because the array did not include the probes.
    - Notice that `cg03`, `cg06`, `cg11`, `cg15`, and `cg18` were included among the surrogate weights, but were not in our methylation data.

```{r, echo = FALSE}
my_surrogate$methyl |>
  data.frame() |>
  rownames_to_column(var = "probe") |>
  mutate(
    across(starts_with("samp"),
           ~ case_when(
             is.na(.x) & probe %in% c("cg02", "cg07") ~ "<font color='blue'><b>NA</b></font>",
             is.na(.x) & probe %in% c("cg03", "cg06", "cg11", "cg15", "cg18") ~ "<font color='red'><b>NA</b></font>",
             .default = format(round(.x, 3), nsmall = 3)
           )
    )
  ) |>
  column_to_rownames(var = "probe") |>
  knitr::kable(format = "html",
               escape = FALSE,
               row.names = TRUE,
               align = rep("c", 5))
```
## Checking for Missing Values

Next, we can check our new methylation matrix for missing values. This seems silly with our small example, but can be much more helpful when working with real data.

```{r}
missing <- methyl_miss(methyl_surro = my_surrogate)
print(missing)
```

Notice that the object created by `methyl_miss()` contains an element for each type of missing data:

- For missing observations (`missing_obs`), the function returns a named vector of the proportion of samples that are missing each of the probes.
- For missing probes (`missing_probes`), the function returns a character vector of probes that are missing for all samples.

## Imputing Missing Observations

Now that we know about our missing values, we'll begin addressing them by imputing missing observations with `impute_obs()`. Required inputs to this function are the `methyl_surro` object, a character string to denote the method (currently available options are "mean" and "median"). The optional `min_nonmiss_prop` option can be specified to impute observations for only those probes that are above the specified proportion of non-missing values.

```{r}
my_surrogate <- impute_obs(methyl_surro = my_surrogate,
                           method = "mean")
print(my_surrogate)
```

## Filling Missing Probes from Reference Data

Now that we've imputed missing observations, we need to account for missing probes (i.e., those that are missing for all samples). For this, we'll need to draw on a reference panel (`ref_vec_mean`) to fill in values with the `reference_fill()` function.

```{r}
data("ref_vec_mean")
print(ref_vec_mean)
```

We'll need to specify our `methyl_surro` object, the `reference` object (requires named vector, matrix, or data frame), and the `type` of values to fill. Since we've already accounted for missing observations, we'll use `type = "probes"` to fill missing probes. Alternatively, the function can be used to fill missing observations (`type = "obs"`), or missing observations and probes simultaneously (`type = "all"`).

```{r}
my_surrogate <- reference_fill(methyl_surro = my_surrogate,
                               reference = ref_vec_mean,
                               type = "probes")
print(my_surrogate)
```

# Estimating Predicted Surrogate Values

We're finally ready to generate predicted values! To do this, we'll use the `surro_calc()` function. We once again need to specify `methyl_surro` object, as well as the `transform` method. In this case, our surrogate weights were created using an elastic net regression model with the Gaussian link function, so we'll use `transform = "linear"`. For surrogate weights developed using the logit link function, predicted probabilities can be estimated with `transform = "probability"`. For those developed using the Poisson link function, predicted counts can be generated with `transform = "count"`.

```{r}
estimates <- surro_calc(methyl_surro = my_surrogate,
                        transform = "linear")
print(estimates)
```

# Piping Commands Together

To simplify the process of estimating predicted surrogate values, we can pipe the commands in a single string to clean up our code.

```{r}
estimates <- beta_matrix_miss |>
  surro_set(weights = wts_vec_lin, intercept = "Intercept") |>
  impute_obs(method = "mean") |>
  reference_fill(reference = ref_vec_mean, type = "probes") |>
  surro_calc(transform = "linear")
print(estimates)
```


