---
title: "Getting Started with MethylSurroGetR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with MethylSurroGetR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

DNA methylation surrogates are statistical models that predict biological or clinical outcomes using methylation values from specific CpG sites. **MethylSurroGetR** provides a comprehensive toolkit for working with these surrogates, handling common challenges like missing data and providing flexible calculation methods.

This vignette demonstrates the main features of the package using example data.

## Installation

```{r setup}
library(MethylSurroGetR)
```

# Basic Workflow

The typical workflow involves four main steps:

1. **Create a surrogate object** with `surro_set()`
2. **Handle missing probes** with `reference_fill()`
3. **Impute missing observations** with `impute_obs()` (if needed)
4. **Calculate predictions** with `surro_calc()`

Let's walk through each step with examples.

# Working with Complete Data

## Step 1: Load Your Data

The package includes sample datasets for demonstration:

```{r load-data}
# Load example methylation data (beta values)
data(beta_matrix_comp)

# Load surrogate weights
data(wts_df)

# Load reference values for missing probes
data(ref_df)

# Preview the data
head(beta_matrix_comp)
```

The methylation matrix should have:
- CpG sites as **rows** (with CpG names as row names)
- Samples as **columns**
- Beta values (0-1 scale) as entries

## Step 2: Create a Surrogate Object

Use `surro_set()` to combine your methylation data with surrogate weights:

```{r surro-set}
# Extract weights as a named vector
wts_vec <- setNames(wts_df$wt_lin, rownames(wts_df))

# Create surrogate object
my_surro <- surro_set(
  methyl = beta_matrix_comp,
  weights = wts_vec,
  intercept = "Intercept"
)

# View the structure
str(my_surro)
```

**Key points:**
- `weights` should be a named numeric vector with CpG names
- `intercept` specifies which weight element is the intercept term
- The function automatically aligns probes and adds missing ones as NA

## Step 3: Fill Missing Probes

If your data is missing probes needed by the surrogate, use reference values:

```{r reference-fill}
# Extract reference values
ref_vec <- setNames(ref_df$mean, rownames(ref_df))

# Fill missing probes with reference values
my_surro <- reference_fill(
  methyl_surro = my_surro,
  reference = ref_vec,
  type = "probes"  # Only fill completely missing probes
)
```

**Fill types:**

- `"probes"` - Only fill completely missing probes (recommended)
- `"obs"` - Only fill individual missing observations
- `"all"` - Fill both missing probes and observations

## Step 4: Calculate Predictions

Now calculate the surrogate predictions:

```{r surro-calc}
# Calculate linear predictions
predictions <- surro_calc(
  methyl_surro = my_surro,
  transform = "linear"
)

print(predictions)
```

**Available transformations:**
- `"linear"` - No transformation (default)
- `"probability"` - Logistic transformation (1 / (1 + exp(-x)))
- `"count"` - Exponential transformation (exp(x))

# Handling Missing Data

## Checking for Missing Data

Use `methyl_miss()` to assess missing data patterns:

```{r missing-check}
# Load data with missing values
data(beta_matrix_miss)

# Create surrogate object
surro_miss <- surro_set(
  methyl = beta_matrix_miss,
  weights = wts_vec,
  intercept = "Intercept"
)

# Analyze missing data
missing_info <- methyl_miss(surro_miss)
print(missing_info)
```

This shows:

- Number of complete probes
- Probes with partial missingness
- Completely missing probes
- Overall missing data rate

## Imputing Missing Observations

For probes with partial missingness, use `impute_obs()`:

```{r impute-obs}
# Impute using row means
surro_imputed <- impute_obs(
  methyl_surro = surro_miss,
  method = "mean",
  min_nonmiss_prop = 0.5,  # Only impute if ≥50% observed
  verbose = TRUE
)
```

**Parameters:**

- `method` - "mean" or "median" imputation
- `min_nonmiss_prop` - Minimum proportion of non-missing values required (0-1)
- `return_stats` - Get detailed imputation statistics

## Complete Workflow with Missing Data

```{r complete-workflow}
# Start with missing data
data(beta_matrix_miss)
data(wts_df)
data(ref_df)

# Create vectors
wts_vec <- setNames(wts_df$wt_lin, rownames(wts_df))
ref_vec <- setNames(ref_df$mean, rownames(ref_df))

# Full pipeline
predictions_miss <- beta_matrix_miss |>
  surro_set(weights = wts_vec, intercept = "Intercept") |>
  reference_fill(reference = ref_vec, type = "probes") |>
  impute_obs(method = "mean", min_nonmiss_prop = 0.5) |>
  surro_calc(transform = "linear")

print(predictions_miss)
```

# Different Surrogate Types

## Linear Surrogates

For continuous outcomes (e.g., estimated white blood cell counts):

```{r linear-example}
wts_linear <- setNames(wts_df$wt_lin, rownames(wts_df))

predictions_linear <- beta_matrix_comp |>
  surro_set(weights = wts_linear, intercept = "Intercept") |>
  reference_fill(reference = ref_vec, type = "probes") |>
  surro_calc(transform = "linear")

print(predictions_linear)
```

## Probability Surrogates

For binary outcomes (e.g., disease risk):

```{r probability-example}
wts_prob <- setNames(wts_df$wt_prb, rownames(wts_df))

predictions_prob <- beta_matrix_comp |>
  surro_set(weights = wts_prob, intercept = "Intercept") |>
  reference_fill(reference = ref_vec, type = "probes") |>
  surro_calc(transform = "probability")

print(predictions_prob)
```

## Count Surrogates

For count outcomes (e.g., cell counts):

```{r count-example}
wts_count <- setNames(wts_df$wt_cnt, rownames(wts_df))

predictions_count <- beta_matrix_comp |>
  surro_set(weights = wts_count, intercept = "Intercept") |>
  reference_fill(reference = ref_vec, type = "probes") |>
  surro_calc(transform = "count")

print(predictions_count)
```

# Working with M-values

Some analyses use M-values instead of beta values. The package provides conversion functions:

```{r m-values}
# Load M-value data
data(mval_matrix_comp)

# Convert M-values to beta values
beta_from_m <- convert_m_to_beta(mval_matrix_comp)

# Or convert beta values to M-values
m_from_beta <- convert_beta_to_m(beta_matrix_comp)

# Verify round-trip conversion
all.equal(beta_matrix_comp, 
          convert_m_to_beta(convert_beta_to_m(beta_matrix_comp)))
```

# Advanced Features

## Getting Diagnostic Information

```{r diagnostics}
# Get detailed diagnostics
results <- surro_calc(
  my_surro,
  transform = "linear",
  return_diagnostics = TRUE
)

# View diagnostics
str(results$diagnostics)
```

## Getting Imputation Statistics

```{r imputation-stats}
# Get detailed imputation statistics
surro_with_stats <- impute_obs(
  surro_miss,
  method = "mean",
  return_stats = TRUE
)

# View statistics
print(surro_with_stats$imputation_stats)
```

## Verbose Output

For detailed information during processing:

```{r verbose, eval=FALSE}
# Verbose mode for debugging
predictions_verbose <- surro_calc(
  my_surro,
  transform = "linear",
  verbose = TRUE
)
```

# Best Practices

1. **Always check for missing data** using `methyl_miss()` before calculation

2. **Use appropriate reference values** when filling missing probes (ideally from the same platform and tissue type)

3. **Set reasonable thresholds** for imputation (e.g., `min_nonmiss_prop = 0.5`)

4. **Choose the correct transformation** based on how the surrogate was developed

5. **Validate results** by comparing to known values or expected ranges

6. **Document your workflow** including software versions and parameters

# Common Issues and Solutions

## Issue: "No common probes found"

**Solution:** Ensure your CpG names match between methylation data and weights. Check for:

- Different naming conventions (e.g., "cg12345678" vs "cg12345678_1")
- Leading/trailing whitespace
- Case sensitivity

## Issue: Many samples removed due to missing data

**Solution:**

- Use `impute_obs()` before `surro_calc()`
- Lower `min_nonmiss_prop` threshold if appropriate
- Use reference values to fill missing probes

## Issue: Predictions outside expected range

**Solution:**

- Verify you're using the correct transformation
- Check that input data is on the correct scale (beta vs M-values)
- Ensure reference values are appropriate for your data

# Summary

MethylSurroGetR provides a complete workflow for:

- Organizing methylation data and surrogate weights
- Handling missing probes and observations
- Calculating predictions with various transformations
- Diagnosing and documenting the analysis process

For more information, see the function documentation (`?surro_calc`, etc.) or visit the [package website](https://github.com/jagoode27/MethylSurroGetR).

# Session Information

```{r session-info}
sessionInfo()
```
